---
title: "Module 7"
author: "Flockology"
date: "2024-11-22"
output: html_document
bibliography: BIOL3140.bib

---

```{r setup, include=FALSE}
library(rgbif)
library(tidyverse)
library(parallel)
library(MuMIn)
library(rnoaa)
library(data.table)
library(ggmap)
library(usmap)
library(magick)
library(cowplot)
library(lme4) 
library(car) 
library(data.table) 
library(knitr) 
library(dplyr)
library(purrr)
library(ggplot2)
library(kableExtra) 
library(lubridate)
```
**Introduction:**

Each spring in the northern hemisphere (Nearctic), many neotropical passerines undertake long-distance migrations from Central and South America to temperate regions to forage and breed. Among these migratory birds, species that fly non-stop over the Gulf of Mexico, landing between Texas and Florida, are known as trans-Gulf migrants (TGMs). The timing of their arrival at breeding grounds is crucial for their reproductive success, balancing tradeoffs between early arrival to secure mates and nesting sites and later arrival to align with peak food availability in temperate, seasonal ecosystems @smith2005arrival @newton2008

In the context of climate change, the inability of TGMs to adjust their arrival dates to earlier springs could lead to population declines @both2006climate . How TGMs and other long-distance migrants shift their spring arrival dates in response to warming trends remains an open question  @knudsen2011. To investigate this, studies often rely on traditional methods such as field observations, bird banding, or tracking, which are valuable but limited in their temporal and spatial scales @askeyev2009 @smith2005arrival

The rise of citizen science has transformed the study of bird migration. Programs like eBird, launched in 2002 by the Cornell Laboratory of Ornithology and the National Audubon Society, have amassed over 100 million bird observations from contributors worldwide. eBird data now provide an extensive and taxonomically broad resource, enabling studies of avian migration and phenology across large spatial and temporal scales @sullivan2009ebird

This project leverages eBird species occurrence data from the Global Biodiversity Information Facility (GBIF), accessed via the R package rgbif, alongside meteorological data obtained through NOAA's National Climatic Data Center API using the rnoaa package. The focus is to explore the effects of local weather conditions on the spring arrival times of TGMs in Massachusetts, providing insights into the responses of these migratory species to changing environmental parameters.



**Methods:**
This project aims to analyze the effect of local weather conditions on trans-Gulf migrant (TGM) arrival times in Massachusetts using eBird data and meteorological information. The methods involve downloading occurrence data for five selected TGM species from the Global Biodiversity Information Facility (GBIF) using the rgbif R package, with specific constraints including human observations in Massachusetts between 1990-2019 during spring months. Weather data is compiled from three strategic locations along the assumed migration route (Mobile, AL; Charlotte, NC; and Boston, MA) using NOAA's National Climatic Data Center API through the rnoaa package, allowing for a comprehensive analysis of how local environmental conditions might influence migration phenology.

**Results:**
```{r, Additional Opperations and Analysis 1,eval=FALSE,echo=FALSE,warning=FALSE, message=FALSE, include=FALSE}
curlopts=list(http_version=2) #solved a connection during mclapply().

species <- c("Coccyzus americanus","Chordeiles minor","Antrostomus vociferus", "Hirundo rustica","Archilochus colubris")

gbif_par <- expand_grid(species, y) %>% group_by(species,y) %>% group_split() 

gbif_fun <- function(n){
  occ_data(scientificName = gbif_par[[n]]$species,
           year=gbif_par[[n]]$y,
           month=paste0("3",",","6"),
           limit=5000,
           country="US",
           basisOfRecord = "HUMAN_OBSERVATION",
           stateProvince="Massachusetts")[[2]]
}

system.time(dat_l <-  mclapply(X=1:length(gbif_par),gbif_fun))

dat <- rbindlist(dat.l,fill=T)

#saving data
saveRDS(dat,"massbird.data.RDS")
```

```{r species count vs year,echo=FALSE,warning=FALSE,message=FALSE} 

#year by year totals from occ_data()
dat <- readRDS("massbird.data.RDS")

dat_summary <- dat %>%
  group_by(year, species) %>%
  summarise(count = sum(individualCount, na.rm = TRUE))
```
<center>
```{r, plot, echo=FALSE, fig.cap="**Figure 1.** Occurence of five TGM species in Massachusetts from 1990 to 2019.", warning=FALSE}
dat%>%
  group_by(year,species)%>%
  summarise(count=sum(individualCount,na.rm = T))%>%
  ggplot(aes(x=year,y=count,col=species))+geom_point()+ labs(title="5 TGM species occurance in MA 1990-2019", x="Year", y="Count")
```
**Figure 1** displays the annual occurance of each of the 5 TGM species in Massacusettes over the 20 year period with species depicted by various colors, clearly illustrating the overall increasing TGM trend. 
</center>

```{r,include=FALSE, warning=FALSE}
sta.d <- data.frame(
  longitude = c(-88.25, -81.0, -71.0),
  latitude = c(30.69, 35.22, 42.36),
  name = c("AL", "NC", "MA"))

path <- sta.d %>%
  select(longitude, latitude, name) %>%
  arrange(factor(name, levels = c("AL", "NC", "MA")))

path$name <- factor(path$name, levels = c("AL", "NC", "MA"))

us_map <- map_data("state")
```
<center>
```{r, echo=FALSE,warning=FALSE, fig.cap="**Figure 2** Flight Path of TGM's from Alabama, to North Carolina, to Massachusetts"}
map_path <- ggplot() +
  geom_map(data = us_map, map = us_map, aes(x = long, y = lat, map_id = region), fill = "white", color = "black") +
  geom_point(data = path, aes(x = longitude, y = latitude, color = name), size = 5) +
  geom_path(data = path,aes(x = longitude, y = latitude, group = 1), color = "black", size = 1) +
  geom_label(data = path,  aes(x = longitude, y = latitude, label = name, fill = name),size = 5, color = "white") +  
  scale_color_manual(values = c("AL" = "red", "NC" = "blue", "MA" = "green")) +
  scale_fill_manual(values = c("AL" = "red", "NC" = "blue", "MA" = "green")) +
  labs(title = "Flight Path of TGMs from AL to MA", 
       color = "Station") +
  theme_minimal() + coord_fixed(1.3)
map_path
```
**Figure 2** Migration route for TGMs displaying spatial distribution between stations. 
</center>

<center>
```{r 2 occurancedata, echo=FALSE,warning=FALSE,message=FALSE, fig.cap="**Figure 3** Variation in Arrival Times"}
mc<- dat%>%
  group_by(species, year)%>%
  mutate(date=as.Date(paste0(year,"-",month,"-",day)),
         j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01")))
  )%>%
  group_by(species,year,j.day,date)%>%
  summarise(day.tot=sum(individualCount,na.rm=T))%>%
  group_by(species,year)%>%
  mutate(prop=cumsum(day.tot/sum(day.tot,na.rm = T)))%>%
  filter(year>1999)

mc.pred <- mc%>%
  group_by(year)%>%
 summarize(
   pred=predict(nls(prop~SSlogis(j.day,Asym, xmid, scal)),newdata=data.frame(j.day=min(j.day):max(j.day))),#predict the logistic curve for each species
   j.day=min(j.day):max(j.day),
  )%>%
  left_join(mc%>%dplyr::select(j.day,date)) 

# 25% 
mc.arrive.date <-mc.pred%>%
  group_by(year)%>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

mc.arrive.date%>%
  ggplot(aes(year,j.day))+geom_point()+labs(x="Year", Y="Julian day", title="Variation in Julian day in Which 25% of the Species Arrived Over 20 Year Period.")
```
**Figure 3** Displays the seemingly random pattern of TGM arrival on the various Julian days over the 20 year time period.  
</center>
```{r, 3 min/max, include=FALSE, warning=FALSE}
options(noaakey = "jTfIdOHJgvxYtIgGoFKkISyKebxUwMkz")
sts <- c(
  "GHCND:USW00013894", #Mobile, AL 2k away about 10 days away @200 km/day
  "GHCND:USW00013881", #Charlotte, NC 1000 km away about 6 days away @200 km/day
  "GHCND:USW00014739" #Boston
)

sta.d <- bind_rows( #bind the rows
  lapply(sts,function(x) ncdc_stations(stationid = x)$data ) #use lapply to run through stations
)%>%
  mutate(usmap_transform(.,input_names = c("longitude","latitude"),output_names = c("longitude.1", "latitude.1")))%>% #join transformation of lat/long for projection with usmap
  mutate(name=str_sub(name, -5,-4))%>%#simplify the name column, grab just the state
  mutate(migr.day=c(10,5,0))%>% #so we can look at wind speed 0, 5 or 10 days before arrive in boston
  separate(id,into = c("station.type","id"))%>%
  print()
weather.d <- meteo_pull_monitors(sta.d$id,date_min = "2000-01-01")
```

```{r,  include=FALSE, warning=FALSE}
weather.d <- weather.d%>%
 mutate(year = year(date), date = as.Date(date)) %>%
  group_by(year)%>% #group by year so we can compute julian day
 mutate(j.day=julian(date,origin=as.Date(paste0(unique(year),"-01-01"))), #add julian day
  date2=date,
  wdir.rad=(180-abs(wdf2-180))*pi/180, #radians so we can use a trig function to compute wind vector, scale degrees first to 180 scale to 2x pi and subtract from 180 (wind comes out of a direction)
  wvec=cos(wdir.rad)*-1*awnd # we want a negative value for positive value for 2x pi
  )%>% #store day in new column
  dplyr::select(id,year,date2,j.day,tmin,tmax,wvec)%>% #select the rows we need
  left_join(sta.d%>%select(id,name,migr.day))%>% #add the station id info (ie. name)
  mutate(j.day=j.day+migr.day)#make j.day ahead of BOS according to the migration days away so we can join weather along path
```

```{r, continuing, include=FALSE, warning=FALSE}
mc.arr.weath <- mc.arrive.date%>%
  left_join(weather.d)%>%
  left_join(mc%>%dplyr::select(year,date,j.day))
# twoweeks
weather.wk <-weather.d %>% 
  group_by(year,name) %>% 
  mutate(wk.tmin = frollmean(tmin, n=14,align="right"),
         wk.tmax = frollmean(tmax, n=14,align="right"),
         wk.wvec = frollmean(wvec, n=14,align="right")
         )%>%
  dplyr::select(j.day,date2,name,wk.tmin,wk.tmax,wk.wvec)

mc.arr.weath2 <- mc.arrive.date%>%
  left_join(weather.wk)
```

```{r, warning=FALSE, include=FALSE, 4}
mc.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),mc.arr.weath,na.action = "na.fail")
mc.lmer2 <- lmer(j.day ~ wk.tmin * wk.tmax * wk.wvec + (1|name), mc.arr.weath2, na.action = "na.fail")
```

```{r,5, echo=FALSE, warning=FALSE}
mc.arr.aic <- dredge(mc.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)
mc.arr.aic
```

```{r, 6, warning=FALSE, echo=FALSE}
Anova(mc.lmer)
Anova(mc.lmer2)
best.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),mc.arr.weath2,na.action = "na.fail")
Anova(best.lmer)
```

**Discussion:**
The findings from this study highlight a significant connection between weather patterns and the timing of trans-Gulf migrant (TGM) bird arrivals in Massachusetts between 2000 and 2019. Linear mixed-effects models showed that weather factors, including two-week rolling averages of minimum and maximum temperatures and wind vectors, played a key role in determining the Julian day of arrival for TGMs. Notably, the significant influence of maximum temperature over a two-week period suggests that the timing of migration is more strongly influenced by broader, seasonal weather trends than by daily fluctuations. 

In addition to temperature, wind vectors also showed a notable influence on TGM arrival timing, though their effects were less pronounced compared to temperature. Wind conditions, which can impact the direction and speed of migration, may interact with temperature trends to modify the timing and success of migration @moore1987stopover These findings suggest that TGMs may adjust their migratory strategies in response to varying wind patterns, particularly in the days leading up to their arrival.

Interestingly, the lack of a strong effect from minimum temperature and wind vectors when analyzed independently supports the idea that migration is influenced by cumulative environmental conditions, rather than abrupt changes in weather on a single day. This is consistent with other studies showing that birds rely on long-term weather patterns to optimize their migration timing @smith2005arrival. Moreover, our results contribute to the ongoing debate about how climate change may influence migration patterns. While some studies have suggested that birds may shift their migration timing in response to warmer spring temperatures @both2006climate, our analysis indicates that a more complex set of environmental variables, including wind patterns and two-week temperature trends, may also play a significant role in determining migratory timing.

These results have important implications for understanding the adaptive strategies of migratory birds in a changing climate. Future studies could explore how other environmental factors, such as precipitation and habitat changes, further influence TGM migration. Long-term monitoring efforts, particularly those utilizing citizen science platforms like eBird, could provide valuable data on how birds are responding to both local and regional environmental changes @sullivan2009ebird. Understanding these patterns will be crucial for predicting the impacts of climate change on avian populations, particularly as migratory birds face increasing pressures from both environmental shifts and human activities.

**Author Contributions**
Jax Lanzerotti contributed the introduction, results, discussion, and author contribution sections of the project. Ella provided the methods section and assisted Jax with the results and citations portion of the project. Alex did nothing. 

**References**

